---
title: "ArchR_H4H_installation"
author: "Ronald Wu"
date: "2024-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Build Partition
```{bash installation}
# access build partition
salloc --partition=build -c 1 -t 1:0:0 --mem 2G

# load R 
module load R/4.2.1
R --interactive
```

# R Interactive Mode
Here we install packages interactively
```{bash install package manager}

find.package("renv") # check if already installed
install.packages(renv)

mkdir ArchR/
cd ArchR
```
> Installing package into ‘/usr/local/lib/R/site-library’
(as ‘lib’ is unspecified)
Warning in install.packages("renv") :
  'lib = "/usr/local/lib/R/site-library"' is not writable
Would you like to use a personal library instead? (yes/No/cancel) yes
Would you like to create a personal library
‘/cluster/home/rwu/R/x86_64-pc-linux-gnu-library/4.2’
to install packages into? (yes/No/cancel) yes

## install packages continued
```{r install packages}
# load renv for package management
library(renv)
library(gitcreds)
library(SummarizedExperiment)
# Make sure you're in the ArchR folder
getwd() # "/cluster/home/rwu/ArchR"

# initialize 
renv::init()

# Install BiocManager if it's not already installed
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  renv::install("BiocManager")
}

# install SummarizedExperiment; from bioconductor
## this is done first b/c ArchR installation fails to install this for some reason
renv::install("bioc::SummarizedExperiment")

# install ArchR
  devtools::install_github("GreenleafLab/ArchR", ref="master", repos = BiocManager::repositories())
  

# install extra packages
ArchR::installExtraPackages()
renv::install("pkg::hexbin") # for plotting
renv::install("pkg::harmony") # for batch correction
renv::install("bioc::Seurat")

# install genomes from bioconductor
renv::install("bioc::BSgenome.Hsapiens.UCSC.hg19") # needed to set addArchRGenome to hg19
renv::install("bioc::BSgenome.Hsapiens.UCSC.hg38") # needed to set addArchRGenome to hg38


# snapshot

renv::snapshot()

# check cache paths
renv::paths$cache() # "/cluster/home/rwu/.cache/R/renv/cache/v5/R-4.2/x86_64-pc-linux-gnu"
```


# Test tutorial dataset

```{bash download dataset}
# access build partition
salloc --partition=build -c 1 -t 1:0:0 --mem 2G

# change dir
cd /cluster/home/rwu/ArchR

# load R 
module load R/4.2.1
R --interactive
```

```{r restore renv}
renv::init()
new_path <- "/cluster/home/rwu/ArchR/renv/library/R-4.2/x86_64-pc-linux-gnu/"
.libPaths(c(new_path, .libPaths()))

find.package("ArchR")
find.package("BSgenome.Hsapiens.UCSC.hg38")
```
> .libPaths()
[1] "/cluster/home/rwu/R/x86_64-pc-linux-gnu-library/4.2"
[2] "/usr/local/lib/R/site-library"
[3] "/usr/lib/R/site-library"
[4] "/usr/lib/R/library"


```{r download dataset}
# download tutorial data (0.5 gigs to current working directory)
inputFiles <- getTutorialData("Hematopoiesis")
```

```{bash mv dataset}
mkdir data/
mv -f /cluster/home/rwu/ArchR/HemeFragments/ /cluster/projects/wouterslab/ArchR/data
```

# Load libraries
```{r load libraries}
library(here)
library(ArchR)
```
```{r input}
inputFiles <- c("data/HemeFragments/scATAC_BMMC_R1.fragments.tsv.gz",
                "data/HemeFragments/scATAC_CD34_BMMC_R1.fragments.tsv.gz",
                "data/HemeFragments/scATAC_PBMC_R1.fragments.tsv.gz")
names(inputFiles) <- c("scATAC_BMMC_R1","scATAC_CD34_BMMC_R1", "scATAC_PBMC_R1")
```

```{r creating arrow files}
# set genome to hg19
addArchRGenome("hg19")

# create arrow files (subset, first sample only)
ArrowFiles <- createArrowFiles(
  inputFiles = inputFiles,
  sampleNames = names(inputFiles),
  minTSS = 4, # Dont set this too high because you can always increase later
  minFrags = 1000, 
  addTileMat = TRUE,
  addGeneScoreMat = TRUE
)

ArrowFiles

```