---
title: "ArchR_Extended_Visualizations"
author: "Ronald Wu"
date: "2024-06-26"
output: html_document
---

# Introduction

Although ArchR has built in plotting functions useful for visualizing high dimensional data from ArchR processed scATAC data, additional plotting functionality is often desired. Here we will explore using `scCustomize` and `Seurat` plotting functions to improve visualization
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install Packages
```{r install packages, eval = FALSE}
install.packages("scCustomize")

BiocManager::install(c("ComplexHeatmap", "dittoSeq", "DropletUtils", "Nebulosa"))
```

# Load Libraries
```{r load libraries}
library(here)
library(ArchR) # version 1.0.2

library(scCustomize)
```

# Load ArchRProject
```{r load ArchRProject}

proj <- loadArchRProject(path = "Save-proj")
```
# Extract ArchRProject Data for scCustomize

scCustomize is designed to work primarily on Seurat objects. Thus, ArchRProject objects are not directly compatible with scCustomize and require extraction of their relevent data for use.

```{r extract data}

# ** optional **
# load previously saved GeneScoreMatrix:
if (exists("exprMatrix")) {
  print("exprMatrix already loaded in environment")
} else {
  file_path <- "Terekhanova/AUCell/TerekhanovaExprMatrix.RData"
  if (file.exists(file_path)) {
    print("Loading exprMatrix from directory")
    load(file = file_path)
  } else {
    print("exprMatrix not in environment or directory, loading from ArchRProject object")
    exprMatrix <- getMatrixFromProject(proj, useMatrix = "GeneScoreMatrix")
    
    # Uncomment the following line if you want to save the matrix for future use
    # save(exprMatrix, file = file_path)
  }
}

# Extract GeneScoreMatrix from ArchRProject
# gene_exp_matrix <- getMatrixFromProject(
#   ArchRProj = proj,
#   useMatrix = "GeneScoreMatrix",
#   verbose = FALSE
# )

# add rownames
rownames(exprMatrix) <- rowData(exprMatrix)$name

# Create Seurat Object
seurat_obj <- CreateSeuratObject(counts = assay(exprMatrix),
                                 assay = "GeneScoreMatrix")

# Add cell metadata
cell_metadata <- getCellColData(ArchRProj = proj)
seurat_obj <- AddMetaData(seurat_obj, metadata = as.data.frame(cell_metadata))

# Add dimensionality reduction embeddings
# For example, adding UMAP embeddings
umap_embeddings <- getEmbedding(ArchRProj = proj, 
                                embedding = "UMAP_harmony", 
                                returnDF = FALSE # returns the object instead of df
                                )

###
# 4. Add dimensionality reduction embeddings (updated with error handling)
# Function to safely add embeddings
add_embedding <- function(embedding_name) {
  tryCatch({
    embeddings <- getEmbedding(ArchRProj = proj, embedding = embedding_name)
    
    # Check if embeddings are not empty
    if (ncol(embeddings) > 0 && nrow(embeddings) > 0) {
      # Ensure embeddings match cells in Seurat object
      embeddings <- embeddings[colnames(seurat_obj), ]
      
      seurat_obj[[tolower(embedding_name)]] <- CreateDimReducObject(
        embeddings = as.matrix(embeddings),
        key = paste0(toupper(embedding_name), "_")
      )
      cat(paste("Added", embedding_name, "embeddings.\n"))
    } else {
      cat(paste("Warning:", embedding_name, "embeddings are empty.\n"))
    }
  }, error = function(e) {
    cat(paste("Error adding", embedding_name, "embeddings:", e$message, "\n"))
  })
  
  return(seurat_obj)
}

# Add the UMAP_harmony embedding (previously made in the ArchRProject object proj)
seurat_obj <- add_embedding("UMAP_harmony")

seurat_obj@reductions$umap_harmony

# change orig.ident to harmony clusters
seurat_obj$orig.ident <- seurat_obj$Clusters_harmony

# Switching Identity in Seurat object to one of the meta.data columns: "Clusters_harmony"
Idents(seurat_obj) <- "Clusters_harmony"

# To switch back to orig.ident
Idents(seurat_obj) <- "orig.ident"


```
> Warning: Feature names cannot have underscores ('_'), replacing with dashes ('-')Warning: Feature names cannot have underscores ('_'), replacing with dashes ('-'). 



# scCustomize Plotting functions

## UMAP
```{r UMAP with Seurat}
# show clusters
DimPlot(seurat_obj, reduction = "umap_harmony", group.by = "Clusters_harmony")

DimPlot_scCustom(seurat_object = seurat_obj)

# show samples
DimPlot_scCustom(seurat_object = seurat_obj,
                 group.by = "Sample")
```

```{r UMAP with scCustomize}

# Use scCustomize functions for plotting
# Example: Create a UMAP plot with custom color palette

FeaturePlot_scCustom(
  seurat_obj = seurat_obj, 
  features = "TSSEnrichment",
  reduction = "umap_harmony",
  colors_use = viridis_plasma_dark_high,
  pt.size = 0.1,
  order = FALSE
)

# Move positive cells to the front of plot; setting order = TRUE
FeaturePlot_scCustom(
  seurat_obj = seurat_obj, 
  features = "TSSEnrichment",
  reduction = "umap_harmony",
  colors_use = viridis_plasma_dark_high,
  pt.size = 0.1,
  order = TRUE
)

```

## UMAP with Cluster_Highlight_Plot for Clusters
```{r UMAP Cluster Highlight}

# Try plotting discrete meta data: Clusters_harmony

## 1. Switching Identity in Seurat object to one of the meta.data columns: "Clusters_harmony"
Idents(seurat_obj) <- "Clusters_harmony"

## 2. Plot

Cluster_Highlight_Plot(seurat_object = seurat_obj, cluster_name = "C1", highlight_color = "navy",
    background_color = "lightgray")

Cluster_Highlight_Plot(seurat_object = seurat_obj, cluster_name = "C20", highlight_color = "forestgreen",
    background_color = "lightgray")

## 3. Show multiple identities

Cluster_Highlight_Plot(
  seurat_object = seurat_obj,
  cluster_name = c("C1", "C20"),
  highlight_color = c("navy", "forestgreen"),
  background_color = "lightgray"
)

```
## FeaturePlots
```{r feature plots}
FeaturePlot_scCustom(seurat_object = seurat_obj, features = "Neftel_MES", alpha_exp = 0.75)
FeaturePlot_scCustom(seurat_object = seurat_obj, features = "AUCell_Neftel_MES", alpha_exp = 0.75)

FeaturePlot_scCustom(seurat_object = seurat_obj, features = "Neftel_MES1", alpha_exp = 0.75)
FeaturePlot_scCustom(seurat_object = seurat_obj, features = "AUCell_Neftel_MES1", alpha_exp = 0.75)

FeaturePlot_scCustom(seurat_object = seurat_obj, features = "Neftel_MES2", alpha_exp = 0.75)
FeaturePlot_scCustom(seurat_object = seurat_obj, features = "AUCell_Neftel_MES2", alpha_exp = 0.75)

FeaturePlot_scCustom(seurat_object = seurat_obj, features = "PIMO_pos", alpha_exp = 0.75)
FeaturePlot_scCustom(seurat_object = seurat_obj, features = "AUCell_PIMO_pos", alpha_exp = 0.75)

FeaturePlot_scCustom(seurat_object = seurat_obj, features = "DAEG_UP_24_newThr", alpha_exp = 0.75)
FeaturePlot_scCustom(seurat_object = seurat_obj, features = "AUCell_DAEG_UP_24", alpha_exp = 0.75)

```
## UMAP with Meta_Highlight_plot for any meta data (factor, character, logical)
```{r UMAP Cluster Highlight}
# meta data cols
names(seurat_obj@meta.data)

# highlight
Meta_Highlight_Plot(
  seurat_object = seurat_obj,
  meta_data_column = "PIMO_pos",
  meta_data_highlight = TRUE,
  highlight_color = "firebrick",
  background_color = "lightgray"
)

Meta_Highlight_Plot(
  seurat_object = seurat_obj,
  meta_data_column = "DAEG_UP_24_newThr",
  meta_data_highlight = TRUE,
  highlight_color = "firebrick",
  background_color = "lightgray"
)

# With Seurat 
DimPlot(seurat_obj, reduction = "umap_harmony", group.by = "PIMO_pos")
DimPlot(seurat_obj, reduction = "umap_harmony", group.by = "DAEG_UP_24_newThr")

```
## Highlight cells based on Marker
```{r highlight cells based on marker}

markerGenes  <- c(
    "CD34", #Early Progenitor
    "GATA1", #Erythroid
    "PAX5", "MS4A1", #B-Cell Trajectory
    "CD14", #Monocytes
    "CD3D", "CD8A", "TBX21", "IL7R" #TCells
  )

# Get cell names
MS4A1 <- WhichCells(object = seurat_obj, expression = MS4A1 > 1)
CA9 <- WhichCells(object = seurat_obj, expression = CA9 > 2)
CD34 <- WhichCells(object = seurat_obj, expression = CD34 > 5)

# Make into list
cells <- list(
              CD34 = CD34)

# Plot
Cell_Highlight_Plot(seurat_object = seurat_obj, cells_highlight = cells)

assay(exprMatrix)["CD34",] %>% range()
assay(exprMatrix)["CD34",] %>% hist
assay(exprMatrix)["CD34",] %>% quantile
rm(CD34)
```

# FeatureScatter Plots 
```{r scatter plots}
names(seurat_obj@meta.data)

# Create Plots
FeatureScatter_scCustom(
  seurat_object = seurat_obj,
  feature1 = "AUCell_PIMO_pos",
  feature2 = "AUCell_DAEG_UP_24",
  colors_use = viridis_plasma_dark_high,
  ggplot_default_colors = TRUE,
  split.by = "PIMO_pos",
  group.by = "ident",
  #num_columns = 1,
  pt.size = 1
)
```
Can include correlation 

# PDF compilation
```{r pdf compilation, eval = FALSE}
here::here()
getOutputDirectory(proj)

# save meta data colnames
meta_data <- seurat_obj@meta.data %>% names
meta_data


# plotPDF(u1,u2,u3,u4,u5,
#         name = "Terekhanova-scCustomPlot-UMAP-HarmonyClusters-AUCell-scores.pdf", 
#         ArchRProj = proj, 
#         addDOC = TRUE, 
#         width = 5, height = 5)

# pdf(here::here("Terekhanova","Plots",
#          "2024-07-03_Terekhanova_scCustom_passes_AUCell_all_geneSets_Harmony_UMAP.pdf"))
# # show samples
# DimPlot_scCustom(seurat_object = seurat_obj,
#                  group.by = "Sample")
# 
# # harmony clusters
# DimPlot_scCustom(seurat_object = seurat_obj) +
#   ggtitle("Harmony Clusters")
# 
# Meta_Highlight_Plot(
#   seurat_object = seurat_obj,
#   meta_data_column = "PIMO_pos",
#   meta_data_highlight = TRUE,
#   highlight_color = "firebrick",
#   background_color = "lightgray"
# ) + 
#   ggtitle("PIMO_pos")
# 
# Meta_Highlight_Plot(
#   seurat_object = seurat_obj,
#   meta_data_column = "DAEG_UP_24_newThr",
#   meta_data_highlight = TRUE,
#   highlight_color = "firebrick",
#   background_color = "lightgray"
# ) +
#   ggtitle("DAEG_UP_24_newThr")
# 
# dev.off()


pdf(here::here("Terekhanova","Plots",
         "2024-07-03_Terekhanova_scCustom_passes_AUCell_all_geneSets_Harmony_UMAP.pdf"))
# show samples
DimPlot_scCustom(seurat_object = seurat_obj,
                 group.by = "Sample")

# harmony clusters
DimPlot_scCustom(seurat_object = seurat_obj) +
  ggtitle("Harmony Clusters")

# plot the binary cell assignments from AUCell
for (i in 21:31) {
  z <-
    Meta_Highlight_Plot(
      seurat_object = seurat_obj,
      meta_data_column = meta_data[i],
      meta_data_highlight = TRUE,
      highlight_color = "firebrick",
      background_color = "lightgray"
    ) +
    ggtitle(meta_data[i])
  
  print(z)
}

# plot the AUCell scores
for (i in 32:41) {

z <- FeaturePlot_scCustom(seurat_object = seurat_obj, features = meta_data[i], alpha_exp = 0.75) +
        ggtitle(paste0("AUCell score: ", meta_data[i]))

  print(z)
}


dev.off()
```



