---
title: "AUCell_Terekhanova"
author: "Ronald Wu"
date: "2024-05-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install Packages
```{r install packages, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("AUCell")
install.packages("NMF")
```

# Load Libraries
```{r load libraries}
library(AUCell)
library(ArchR) # version 1.0.2
library(here)
library(tidyverse)
library(NMF)
```

# Load ArchRProject

We load the previous ArchRProject we've been working on. See `ArchR_Terekhanova.Rmd` for more details.

```{r load ArchRProject, eval = FALSE}
# load ArchRProject from Terekhanova/
proj <- loadArchRProject(path = "Terekhanova")
```

# AUCell

## Load scATAC ArchR object and extract GeneScoreMatrix
```{r AUCell, eval = FALSE}

# create a folder for AUCell results
dir.create("Terekhanova/AUCell/")

# load exprMatrix if not already in memory; otherwise create it
if (!exists("exprMatrix")) {
  load(file = "Terekhanova/AUCell/TerekhanovaExprMatrix.RData") # 1.1 gb
} else{
  # load expression matrix (from ArchRProject object)
  ## Access the GeneScoreMatrix as SummarizedExperiment
  exprMatrix <- getMatrixFromProject(proj,
                                     useMatrix = "GeneScoreMatrix")
  
  # save the matrix for easier load in the future
  save(exprMatrix, file = "Terekhanova/AUCell/TerekhanovaExprMatrix.RData") # 1.1 gb
  
}

# access the sparse dgCMatrix in the SummarizedExperiment 
assayNames(exprMatrix) # GeneScoreMatrix
assay(exprMatrix) %>% dim # 24919 30187


# add gene names to the rownames slot 
rownames(exprMatrix) <- rowData(exprMatrix)$name
# rownames(assay(exprMatrix)) <- exprMatrix@elementMetadata$name
```

Since the `getMatrixFromProject()` function from ArchR returns a SummarizedExperiment object containing a sparse `dgCMatrix`, we can do without conversion to a dgCMatrix.  

## Load gene sets

AUCell has implemented functions to utilize signatures obtained using the `GSEAbase` package. We will skip using those functions, since our external gene sets of interest are `.txt` files. They have been moved to the `data/Signatures/` folder.

```{r load gene sets}
# load DAEG and Neftel signatures saved as previously saved as .txt

## get paths
txt_files <- list.files(path = here("data","Signatures"), pattern = "\\.txt$", full.names = TRUE)

## read the .txt files into a list
geneSets <- lapply(txt_files, read_lines)

## name the elements of the list based on the original file name (without path and extension)
names(geneSets) <- tools::file_path_sans_ext(basename(txt_files))

## *** Load additional gene sets (not txt files) *** #
PIMO_pos <- read_csv(here("data","Signatures","pimo_sig",
              "pos_vs_neg","pimo_pos_rankings.csv")) %>%
  pull(names)
  
PIMO_pos %>% length # 267
###########

## append additional gene sets to list
geneSets <- append(geneSets, list(PIMO_pos = PIMO_pos))



## add the gene set size to the names (req GSEABase obj)
# geneSets <- setGeneSetNames(geneSets, newNames=paste(names(geneSets), " (", nGenes(geneSets) ,"g)", sep=""))

# add hypoxia signatures:

# function that checks if genes in the geneSets are in the GeneScoreMatrix

calculate_overlap <- function(se, geneSets) {
  # extract gene symbols from the SummarizedExperiment
  se_genes <- rownames(se)
  # make data frame to store info
  results <- data.frame(
    "Gene_Set" = character(),
    "Set_Size" = integer(),
    "Overlap" = integer(),
    "Percentage_OL" = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Loop through each gene set
  for (set_name in names(geneSets)) {
    gene_set <- geneSets[[set_name]]
    
    ## Calculate overlap
    overlap <- length(intersect(se_genes, gene_set))
    set_size <- length(gene_set)
    percentage <- (overlap / set_size) * 100
    
    ## Add results to the data frame
    results <- rbind(
      results,
      data.frame(
        "Gene_Set" = set_name,
        "Set_Size" = set_size,
        "Overlap" = overlap,
        "Percentage_OL" = percentage,
        stringsAsFactors = FALSE
      )
    )
  }
  return(results)
}

calculate_overlap(exprMatrix, geneSets) # check if geneset genes are in the matrix

# additional test of custom genes
testSets <- list(a = c("CA9","GLUT1","SLC2A1"))
calculate_overlap(exprMatrix, testSets)
```
Based on the calculated overlap table, all the genes for each geneset except for 1 or 2 are present in the scATAC data, which is good. The size of the genesets are small however.

# score gene signatures
```{r score gene signature, eval = FALSE}
# if cells_AUC if not already in memory; load it or create it
if (!exists("cells_AUC")) {
  if (file.exists("Terekhanova/AUCell/cells_AUC.RData")) {
    load(file = "Terekhanova/AUCell/cells_AUC.RData") # 2.2 mb
  }
  else{
    print("file not found. creating cells rankings and cell AUC")
    
    # calculates both cell ranking and enrichment
    # cells_AUC <- AUCell_run(exprMatrix, geneSets)
    
    # alternative: calculate cell ranking and enrichment seperately
    cell_rankings <- AUCell_buildRankings(exprMatrix, plotStats = FALSE)
    cells_AUC <- AUCell_calcAUC(geneSets, cell_rankings)
    
    # save cells_AUC
    save(cells_AUC, file = "Terekhanova/AUCell/cells_AUC.RData")
  }

cell_rankings
cells_AUC

# check # of expressed/detected genes in dataset
plotGeneCount(assay(exprMatrix), 
              plotStats = TRUE, verbose = TRUE)
```
Quantiles with the number of genes detected by cell 
>     min      1%      5%     10%     50%    100% 
  362.0   759.0  1338.3  2053.0  7396.0 21554.0 
  
> Genes in the gene sets NOT available in the dataset: 
	Neftel_AC: 	1 (3% of 39)
	Neftel_MES: 	2 (2% of 95)
	Neftel_MES1: 	1 (2% of 50)
	Neftel_MES2: 	1 (2% of 50)
	Neftel_NPC: 	4 (4% of 89)
	Neftel_NPC1: 	2 (4% of 50)
	Neftel_NPC2: 	3 (6% of 50)
	Neftel_OPC: 	1 (2% of 50)
	PIMO_pos: 	49 (18% of 267)

As mentioned in the vignette, the top 5% of genes in the rankings are used by default as the threshold to calculate AUC. We can adjust the `aucMaxRank` parameter in the `AUCell_run()` or `AUCell_calcAUC()` function depending on if we think our dataset expresses more genes. Due to the sparsity in the scATAC data, I think we will leave it to speed up analysis.

For each gene set, AUCell asks the question, are the genes in the gene set detected in the top 5% (default) genes ranked in a given cell.

## Explore thresholds

Depending on the type of distribution for each gene set, several thresholds are calculated and assigned by default. We can visually inspect these thresholds by plotting their histograms. Manual intervention can be done to reassign a threshold if a given distribution is ambiguous.

```{r thresholds, eval = FALSE}
set.seed(42)
par(mfrow=c(3,3))

# plot all AUC histograms (for all geneSets) and calculates likely thresholds
pdf(here("Terekhanova","Plots","2024-06-07_Terekhanova_AUCell_cells_assignment.pdf"))
cells_assignment <- AUCell_exploreThresholds(cells_AUC, 
                                             plotHist = TRUE,
                                             assign = TRUE)

dev.off()

warningMsg <- sapply(cells_assignment, function(x) x$aucThr$comment)
warningMsg[which(warningMsg!="")]

# look at possible thresholds for (DAEG_UP_24)
cells_assignment$DAEG_UP_24$aucThr$thresholds
cells_assignment$Neftel_MES2$aucThr$thresholds

# look at auto-selected thresholds for (DAEG_UP_24)
cells_assignment$DAEG_UP_24$aucThr$selected # minimumDens
# of cells assigned 
cells_assignment$DAEG_UP_24$assignment %>% length() # 23353
```

## manual threshold setting

Let's inspect the distribution of AUC for gene sets of interest. For DAEG_UP_24, the auto-selected AUC threshold minimumDens was 0.01381947, representing the inflecting point of the density curve. 

```{r manual thresholds}
# plot AUC histogram for DAEG_UP_24 gene set: 
geneSetName <- rownames(cells_AUC)[grep("DAEG_UP_24", rownames(cells_AUC))]
AUCell_plotHist(cells_AUC[geneSetName,], aucThr= 0.01381947,
                title = "old threshold")
abline(v= 0.01381947)

# set new threshold
newThr = 0.10400131 # L_k2 threshold

# get barcodes for cells that pass new threshold
newSelectedCells <- names(which(getAUC(cells_AUC)[geneSetName,]>newThr))
length(newSelectedCells) # 1448

# get barcodes for cells for previous default threshold
oldSelectedCells <- cells_assignment$DAEG_UP_24$assignment
length(oldSelectedCells) # 23353
```

When looking at the density curve, there is no obvious inflection point. There is also no obvious bimodal distribution. We do expect hypoxic cells to be a gradient rather than a positive vs negative bimodal distribution.

If we use this default threshold, of the 30187 cells in the dataset, majority of the cells (23353; 77%) would be classified as positive for this signature. That seems high. We might need to manually set threshold for DAEG_UP_24.

## AUCell histogram
```{r save AUCell histogram, eval = FALSE}

# save PDF for AUCell histogram (all geneSets)
pdf(here("Terekhanova", "Plots",
         "2024-06-04_Terekhanova_AUCell_default_aucThr_plotHist.pdf"))
for (i in names(cells_AUC)) {    
  aucThr <- as.numeric(cells_assignment[[i]]$aucThr$selected)
                    
  AUCell_plotHist(cells_AUC[i, ], aucThr = aucThr)
  abline(v = aucThr) # vertical line at threshold
  text( # prints threshold on left side of plot
    x = 0.15,
    y = 500,
    labels = print(paste0("aucThr: ", round(aucThr,5),"\n",
           "CellsAssigned: ", length(cells_assignment[[i]]$assignment)
           )),
    cex = 1,
    col = "red",
    adj = 1
  )
}
dev.off()

pdf(here("Terekhanova", "Plots",
         "2024-06-04_Terekhanova_AUCell_new_aucThr_plotHist.pdf"))
  AUCell_plotHist(cells_AUC[geneSetName,], aucThr= newThr,
                  title = "DAEG_UP_24\nnew threshold (L_k2)")
  abline(v= newThr)
  text( # prints threshold on left side of plot
    x = 0.15,
    y = 500,
    labels = print(paste0("aucThr: ", round(newThr,5),"\n",
           "CellsAssigned: ", length(newSelectedCells)
           )),
    cex = 1,
    col = "red",
    adj = 1
  )
dev.off()

```
# Exploring the cell-assignment (table & heatmap)

Given that we've assigned the cells (their barcodes) basesd on their AUC for each given gene set, we can visualize this through a binary incidence matrix via a heatmap.
```{r tabulate cell assignments}
# cells for each patient
table(proj$Sample)

# hypoxic cells for each patient
table(proj$Sample, proj$DAEG_UP_24_newThr)
table(proj$Sample, proj$PIMO_pos)

# Neftel states for each patient
table(proj$Sample, proj$Neftel_AC)
table(proj$Sample, proj$Neftel_MES)
table(proj$Sample, proj$Neftel_OPC)
table(proj$Sample, proj$Neftel_NPC)

# Neftel_MES and hypoxic 
table(proj$Neftel_MES)
table(proj$Neftel_MES, proj$DAEG_UP_24_newThr)
table(proj$Neftel_MES, proj$PIMO_pos)
```

We can also use getGroupSummary() to summarize cellColData 
```{r group summary}
# access cellColData
proj@cellColData

# summary example
getGroupSummary(
  ArchRProj = proj,
  groupBy = "PIMO_pos",
  select = "TSSEnrichment", # must be numeric column
  summary = "median",
  removeNA = TRUE
)
```

```{r cell assignment table and heatmap}
# grabs cell barcodes of cells assigned to given geneSet according to default thresholds
cellsAssigned <- lapply(cells_assignment, function(x) x$assignment)

# add barcodes of new threshold for DAEG_UP_24
cellsAssigned[["DAEG_UP_24_newThr"]] <- newSelectedCells

# converts list to a tibble
assignmentTable <- cellsAssigned %>%
  enframe(name = "geneSet", value = "cell") %>%
  unnest(cell)
  #pivot_longer(cols = -geneSet, names_to = "variable", values_to = "cell")

head(assignmentTable)

# Convert into an incidence matrix 
assignmentTable <- as.data.frame(assignmentTable)
assignmentMat <- table(assignmentTable[,"geneSet"], assignmentTable[,"cell"])

# quick inspection
assignmentMat[,1:2]

# plot incidence matrix as heatmap:
library(NMF)
set.seed(42)

# draw of subset matrix (fast)
## subset 100 cells
miniAssigMat <- assignmentMat[,sample(1:ncol(assignmentMat),100)]

aheatmap(
  miniAssigMat,
  scale = "none",
  color = "black",
  legend = FALSE,
  main = "Assignment Matrix AUCell: Terekhanova GBM (n=4)\n Subset: 100 cells",
)

## subset 1000 cells
miniAssigMat1000 <- assignmentMat[,sample(1:ncol(assignmentMat),1000)]
aheatmap(
  miniAssigMat1000,
  scale = "none",
  color = "black",
  legend = FALSE,
  main = "Assignment Matrix AUCell: Terekhanova GBM (n=4)\n Subset: 1000 cells",
)

# draw full matrix (very slow; takes too long)
# aheatmap(assignmentMat, scale="none", color="black", legend=FALSE)


```
## export PDF of assignment heatmap
```{r export PDF assignment heatmap, eval = FALSE}
pdf(here("Terekhanova", "Plots",
         "2024-06-07_Terekhanova_AUCell_miniAssigMat_heatmap.pdf"))

aheatmap(
  miniAssigMat,
  scale = "none",
  color = "black",
  legend = FALSE,
  main = "Assignment Matrix AUCell: Terekhanova GBM (n=4)\n Subset: 100 cells",
)

aheatmap(
  miniAssigMat1000,
  scale = "none",
  color = "black",
  legend = FALSE,
  main = "Assignment Matrix AUCell: Terekhanova GBM (n=4)\n Subset: 1000 cells",
)
dev.off()

```

# Colorize UMAPs with AUCell assigned cell types/states

## Add AUCell assignments to ArchRProject object

One way to plot the AUCell assignments is to append additional info into the CellColData slot in the ArchRProject object. We can then use the `ArchR::plotEmbedding()` function to plot the UMAP with additional parameters, such as highlighting the cells (or their barcodes) that have been assigned to a given gene set. These assignments have been determined previously (see above) through `AUCell` and pass a specific AUC score threshold. 

```{r AUCell to ArchRProject}
# ArchRProject Object
proj

# examine CellColData in ArchRProject Object
getCellColData(proj)
exprMatrix@colData
getCellNames(proj)

# name of gene sets to add as new columns
new_cols <- cells_AUC@NAMES

# determine membership of each barcode for each assigned gene Set
new_cols_list <- list() # initialize list
  # loop through each gene set
for (i in names(cells_AUC)) {
  new_cols_list[[i]] <- cells_assignment[[i]]$assignment # extract assigned barcodes
  # creates new column in cellColData
  proj@cellColData[[i]] <-
    ifelse(getCellNames(proj) %in% new_cols_list[[i]], TRUE, FALSE) # determine membership
}

# manual add new columns to CellColData: for DAEG_UP_24 new threshold 
proj$DAEG_UP_24_newThr <- ifelse(getCellNames(proj) %in% newSelectedCells, TRUE, FALSE)

# manual add new columns to CellColData: for PIMO_pos
proj$PIMO_pos <- ifelse(getCellNames(proj) %in% cells_assignment$PIMO_pos$assignment, 
                        TRUE, FALSE)

# test if correct
cells_assignment$Neftel_MES2$assignment %>% length # 822
proj@cellColData$Neftel_MES2 %>% sum() # 822



#### re-write above to use addCellColData()
## add numeric AUC scores to the cellColData
cells_AUC

auc_matrix <- assay(cells_AUC)

auc_matrix[1:10,1:10] %>% View
auc_matrix["Neftel_AC",1:10]

projTest <- proj
for (set_name in rownames(auc_matrix)) {
  projTest <- addCellColData(
    ArchRProj = projTest,
    data = auc_matrix[set_name, ],
    name = paste0("AUCell_", set_name),
    cells = colnames(auc_matrix)
  )
}

projTest@cellColData
```

## plot UMAP
```{r colorize umaps}
# ArchRProject object
proj

proj@embeddings #

# UMAP with Harmony embeddings
set.seed(42)
## by sample
p1 <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "cellColData", 
    name = "Sample", 
    embedding = "UMAP_harmony",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)
## use hightlightCells param to highlight cells that pass old threshold for DAEG_24_UP
p2 <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "cellColData", 
    name = "Sample", 
    embedding = "UMAP_harmony",
    highlightCells = oldSelectedCells, # barcodes of DAEG_24_UP w/ old thr
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)
## highlight cells that pass new threshold for DAEG_24_UP
p3 <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "cellColData", 
    name = "Sample", 
    embedding = "UMAP_harmony",
    highlightCells = newSelectedCells, # barcodes of DAEG_24_UP w/ newThr
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

# highlight DAEG_24_UP assigned cells without using highlightCells
  ## add a new column in cellColData showing cell assignment 
  ## From AUCell for DAEG_24_UP new threshold (manual)
proj$DAEG_24_UP <- ifelse(getCellNames(proj) %in% newSelectedCells, "yes", "no")

p4 <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "cellColData", 
    name = "DAEG_UP_24_newThr", 
    embedding = "UMAP_harmony",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)


exprMatrix@colData
getCellNames(proj)
getCellColData(proj)



# plot AUCell scores and binary assignments
p5 <- plotEmbedding(
    ArchRProj = projTest, 
    colorBy = "cellColData", 
    name = "AUCell_DAEG_UP_24", 
    embedding = "UMAP_harmony",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

# plot
p1
p2
p3
p4
p5

# check if plots are correct
cells_assignment$PIMO_pos$aucThr$selected # 0.05853127
(projTest$AUCell_PIMO_pos > 0.05853127) %>% sum # 119
projTest$PIMO_pos %>% sum # 119
## do the mean scores their groupings
getGroupSummary(
  ArchRProj = projTest,
  groupBy = "PIMO_pos",
  select = "AUCell_PIMO_pos", # must be numeric column
  summary = "mean",
  removeNA = TRUE
)


```

```{r export pdf, eval=FALSE}
# export pdf
pdf(here("Terekhanova", "Plots",
         "2024-06-19_Terekhanova_passes_AUCell_DAEG_UP_24_newThr_Harmony_cluster_UMAP.pdf"))
p4
dev.off()
```

# Plot Genomic Tracks
Here we use the plotting function from ArchR to show the genomic tracks for genes of interest. Namely, we examine the loci for one of our genesets used in the AUCell analysis. The `ArchR::plotBrowserTrack()` function allows separation of groups of cells and averaging of the ATAC signal based on `CellColData`. 

## With AUCell cell assignments
```{r plot genomic tracks}
# marker genes to plot
geneSets
geneSets[["DAEG_UP_24"]]

# potential groupings for cells
proj@cellColData

# plot tracks at genes for: DAEG_UP_24
# group cells by: harmony clusters
b1 <- plotBrowserTrack(
    ArchRProj = proj, 
    groupBy = "Clusters_harmony", 
    geneSymbol = geneSets[["DAEG_UP_24"]], 
    upstream = 50000,
    downstream = 50000
)

# plot tracks at genes for: DAEG_UP_24
# group cells by: AUCell assignments
## eg. PIMO_pos


# note: groupBy the columns in CellColData; doesn't accept logical
proj$PIMO_pos <- as.factor(proj$PIMO_pos) # convert from logical to factor
b2 <- plotBrowserTrack(
    ArchRProj = proj, 
    groupBy = "PIMO_pos", 
    geneSymbol = geneSets[["DAEG_UP_24"]], 
    upstream = 50000,
    downstream = 50000
)


# plot track for specific genes:
##
grid::grid.newpage()
grid::grid.draw(b2$CA9)

# include scTrack; shows binarized scATAC signal for each cell (100 cells default)
b3 <- plotBrowserTrack(
    ArchRProj = proj, 
    groupBy = "PIMO_pos", 
    geneSymbol = c("HK2","CA9","ERO1A","LOX"),
    #useMatrix = "GeneScoreMatrix",
    plotSummary = c("bulkTrack", "featureTrack","scTrack", "geneTrack"),
    scCellsMax = 100, # default number of cells to show in the binarized scATAC signal track
    upstream = 50000,
    downstream = 50000
)

## DAEG_UP_24_newThr; use matrix to plot gene scores 
proj$DAEG_UP_24_newThr <- as.character(proj$DAEG_UP_24_newThr) # convert from logical to factor

b4 <- plotBrowserTrack(
    ArchRProj = proj, 
    groupBy = "DAEG_UP_24_newThr", 
    geneSymbol = geneSets[["DAEG_UP_24"]],
    #geneSymbol = c("HK2"),
    useMatrix = "GeneScoreMatrix",
    plotSummary = c("bulkTrack", "featureTrack","scTrack", "geneTrack"),
    scCellsMax = 100, # default number of cells to show in the binarized scATAC signal track
    upstream = 50000,
    downstream = 50000
)


grid::grid.newpage()
grid::grid.draw(b4$HK2)
grid::grid.draw(b3$ERO1A) # hypoxia gene NOT part of signature
grid::grid.draw(b3$LOX) # hypoxia gene NOT part of signature

```
## save pdf of genomic track plots

```{r genomic track plots pdf, eval = FALSE}

pdf(here("Terekhanova", "Plots",
          "2024-06-12_Terekhanova_DAEG_UP_24_groupBy_PIMO_pos_Genomic_Track_Plots.pdf"))

for (i in names(b2)){
grid::grid.newpage()
grid::grid.draw(b2[[i]])

}

dev.off()

pdf(here("Terekhanova", "Plots",
         "2024-06-13_Terekhanova_DAEG_UP_24_groupBy_Clusters_harmony_Genomic_Track_Plots.pdf"))

for (i in names(b1)){
grid::grid.newpage()
grid::grid.draw(b1[[i]])

}

dev.off()


pdf(here("Terekhanova", "Plots",
         "2024-06-13_test_Genomic_Track_Plots.pdf"))
grid::grid.newpage()
grid::grid.draw(b3$HK2)
grid::grid.draw(b3$ERO1A) # hypoxia gene NOT part of signature
grid::grid.draw(b3$LOX) # hypoxia gene NOT part of signature
dev.off()


pdf(here("Terekhanova", "Plots",
         "2024-06-19_Terekhanova_DAEG_UP_24_groupBy_DAEG_UP_24_Genomic_Track_Plots.pdf"))

for (i in names(b4)){
grid::grid.newpage()
grid::grid.draw(b4[[i]])

}
dev.off()

```
